---
title: "diabetes_prediction_dataset"
output: 
  pdf_document:
    latex_engine: xelatex
author: ""
date: "`r Sys.Date()`"
---

## Feature Description

### `gender`: Biological Sex  
- **male**: 41%  
- **female**: 59%  
- **other**: 0%  

Gender may influence an individual's susceptibility to diabetes.

---

### `age`: Age  
Age is an important risk factor for diabetes — generally, older individuals have a higher likelihood of developing it.  
The age range in this dataset is **0 to 80 years**.

---

### `hypertension`: Presence of Hypertension  
- `0`: No hypertension  
- `1`: Has hypertension  

Hypertension is a known comorbidity and risk factor for diabetes.

---

### `heart_disease`: Presence of Heart Disease  
- `0`: No heart disease  
- `1`: Has heart disease  

Heart disease is strongly associated with increased diabetes risk.

---

### `smoking_history`: Smoking Behavior  
Smoking is a risk factor for diabetes and can worsen its complications. Categories include:

- `not current`: Previously smoked but not currently  
- `former`: Former smoker, now quit  
- `No Info`: No information available  
- `current`: Currently smoking  
- `never`: Never smoked  
- `ever`: Smoked at some point in the past

---

### `bmi`: Body Mass Index  
BMI is calculated from height and weight to assess body fat. Range: **10.16 to 71.55**

- BMI < 18.5: Underweight  
- 18.5 ≤ BMI < 24.9: Normal  
- 25 ≤ BMI < 29.9: Overweight  
- BMI ≥ 30: Obese

---

### `HbA1c_level`: Hemoglobin A1c  
HbA1c reflects average blood glucose levels over the past **2–3 months**.  
Levels above **6.5%** typically indicate diabetes risk.

---

### `blood_glucose_level`: Blood Glucose Level  
This refers to the concentration of glucose in the blood at a specific time.  
Higher glucose levels are a key indicator of diabetes.

---

### `diabetes`: Target Variable (Diabetes Status)  
- `0`: Does not have diabetes  
- `1`: Has diabetes

Objective1:
Find out which features (such as age, bmi, gender, etc.) may have a significant statistical or pattern relationship with diabetes (values of 0 or 1)

---

Read the Diabetes prediction dataset and view the first five rows of the diabetes_prediction_dataset:


```{r chunk1}
diabetes_dataset = read.csv("./diabetes_prediction_dataset.csv")
head(diabetes_dataset, 5)
```
```{r chunk1.1}
# Check column names and whether they need to be renamed
colnames(diabetes_dataset)

# Replace spaces (if any)
# colnames(diabetes_dataset) <- gsub(" ", "_", colnames(diabetes_dataset))
```
Check if there are exact duplicate rows???
This dataset does not contain unique identifiers, so we do not check for duplicate rows, since each duplicate row is a failed entry.




View dataset information
```{r chunk2}
str(diabetes_dataset)
summary(diabetes_dataset)
```




Check the values of gender and smoking_history. Determine whether to convert to dummy variables (dummy/one-hot) or label encoding.
```{r chunk2.1}
# View all unique values for gender and smoking_history
unique(diabetes_dataset$gender)
unique(diabetes_dataset$smoking_history)

# View frequency distribution in detail
table(diabetes_dataset$gender)
table(diabetes_dataset$smoking_history)
```


Check if there are missing values in the dataset
```{r chunk3}
anyNA(diabetes_dataset)         # any missing values
colSums(is.na(diabetes_dataset))  # num of missing values in each column
```
This data set does not have missing values, so there is no need to fill missing values


```{r chunk2.2}
# convert gender to factor
diabetes_dataset$gender <- factor(diabetes_dataset$gender)

# dummy variables
gender_dummies <- model.matrix(~ gender - 1, data = diabetes_dataset)

# add it back
diabetes_dataset <- cbind(diabetes_dataset, gender_dummies)
head(diabetes_dataset, 5)
```
```{r chunk2.3}
library(dplyr)

# Merge smoking_history categories into more logical groups
diabetes_dataset$smoking_status_grouped <- dplyr::case_when(
  diabetes_dataset$smoking_history == "never" ~ "never",
  diabetes_dataset$smoking_history == "current" ~ "current",
  diabetes_dataset$smoking_history %in% c("former", "not current", "ever") ~ "former",
  TRUE ~ "unknown"
)

# convert gender to factor
diabetes_dataset$smoking_status_grouped <- factor(diabetes_dataset$smoking_status_grouped)

# one-hot encoding
smoking_dummies <- model.matrix(~ smoking_status_grouped - 1, data = diabetes_dataset)

# add it back
diabetes_dataset <- cbind(diabetes_dataset, smoking_dummies)
head(diabetes_dataset, 5)
```


Data Cleaning and Outlier Handling
```{r chunk4}
# Select continuous variables
num_vars <- c("age", "bmi", "HbA1c_level", "blood_glucose_level")

# 2 x 2 
par(mfrow = c(2, 2))  

# boxplot
boxplot(diabetes_dataset$age, 
        main = "Boxplot of Age", 
        col = "lightblue", outline = TRUE)

boxplot(diabetes_dataset$bmi, 
        main = "Boxplot of BMI", 
        col = "lightblue", outline = TRUE)

boxplot(diabetes_dataset$HbA1c_level, 
        main = "Boxplot of HbA1c", 
        col = "lightblue", outline = TRUE)

boxplot(diabetes_dataset$blood_glucose_level, 
        main = "Boxplot of Glucose", 
        col = "lightblue", outline = TRUE)

par(mfrow = c(1, 1))


```
BMI has a large number of outliers, which should be retained or binned.
Blood Glucose Level has an upper bound outlier, but it is likely a true anomaly
HbA1c_level has a few outliers, but not too much
Age distribution is natural, no processing required

```{r chunk5}
# Distribution of age and BMI,HbA1c,Glucose
library(ggplot2)
ggplot(diabetes_dataset, aes(x = age, fill = as.factor(diabetes))) + 
  geom_histogram(binwidth = 5, position = "dodge") +
  labs(title = "Age Distribution by Diabetes Status", fill = "Diabetes")



```
```{r chunk5.1}
ggplot(diabetes_dataset, aes(x = 1:nrow(diabetes_dataset), y = bmi, color = as.factor(diabetes))) +
  geom_point(alpha = 0.4, size = 1) +
  labs(title = "BMI Distribution by Diabetes Status",
       x = "Index", y = "BMI", color = "Diabetes") +
  theme_minimal()
```
```{r chunk5.2}
ggplot(diabetes_dataset, aes(x = 1:nrow(diabetes_dataset), y = HbA1c_level, color = as.factor(diabetes))) +
  geom_point(alpha = 0.4, size = 1) +
  labs(title = "HbA1c Level by Diabetes Status",
       x = "Index", y = "HbA1c Level", color = "Diabetes") +
  theme_minimal()
```
```{r chunk5.3}
ggplot(diabetes_dataset, aes(x = 1:nrow(diabetes_dataset), y = blood_glucose_level, color = as.factor(diabetes))) +
  geom_point(alpha = 0.4, size = 1) +
  labs(title = "Blood Glucose Level by Diabetes Status",
       x = "Index", y = "Glucose Level", color = "Diabetes") +
  theme_minimal()
```
```{r chunk5.4}
ggplot(diabetes_dataset, aes(x = as.factor(hypertension), fill = as.factor(diabetes))) +
  geom_bar(position = "fill") +
  labs(title = "Diabetes Proportion by Hypertension",
       x = "Hypertension (0 = No, 1 = Yes)", y = "Proportion", fill = "Diabetes") +
  theme_minimal()
```

a Class Imbalance
- Observation from all plots: The number of diabetes-positive cases (diabetes = 1) is significantly lower than diabetes-negative cases.
- The dataset shows a strong class imbalance, roughly 1 diabetic case for every 10 non-diabetic cases.

b Age Distribution by Diabetes Status：

Most diabetes cases are concentrated in the 40+ age group. The number of diabetes cases increases steadily from age 40 to 60. This suggests that middle-aged and older adults have a significantly higher risk of diabetes.

c BMI Distribution by Diabetes Status：

Higher BMI is associated with more diabetes cases. There is a clear concentration of diabetic individuals in higher BMI ranges. Indicates that BMI is a strong risk factor for diabetes.

d HbA1c Level by Diabetes Status：

Individuals with HbA1c ≤ 5.0 are almost exclusively non-diabetic. The 6.0–6.5 range shows mixed cases, with relatively fewer diabetics. HbA1c ≥ 7.0 is almost entirely associated with diabetes.

This aligns well with clinical guidelines (HbA1c ≥ 6.5% indicates diabetes). bA1c is a highly predictive variable.

e Blood Glucose Level by Diabetes Status：

Glucose ≤ 125: almost no diabetic cases. Glucose 125–200: mixed, but majority are non-diabetic. Glucose ≥ 200: nearly all are diabetic cases.

This supports medical criteria that blood glucose over 200 mg/dL is a strong indicator of diabetes.

f Diabetes Proportion by Hypertension：

Diabetic proportion is much higher among individuals with hypertension. While only ~10% of non-hypertensive individuals have diabetes, over 25% of hypertensive individuals are diabetic.

Suggests a significant association or comorbidity between hypertension and diabetes.






```{r chunk6}
# Correlation Matrix Heatmap (numeric variables only)
library(corrplot)
corr_matrix <- cor(diabetes_dataset[, num_vars])
corrplot(corr_matrix, method = "color", type = "upper", tl.cex = 0.7)

```
Feature Transformation and Scaling
```{r chunk7}
# Standardize numerical variables
# diabetes_dataset[num_vars] <- scale(diabetes_dataset[num_vars])

```

```{r chunk7}
# how gender affects diabetes
ggplot(diabetes_dataset, aes(x = gender, fill = as.factor(diabetes))) +
  geom_bar(position = "fill") +
  labs(title = "Diabetes Proportion by Gender", y = "Proportion")

# BMI vs Age by diabetes status
ggplot(diabetes_dataset, aes(x = bmi, y = age, color = as.factor(diabetes))) +
  geom_point(alpha = 0.5) +
  labs(title = "BMI vs Age by Diabetes", color = "Diabetes")

# ttset
t.test(age ~ diabetes, data = diabetes_dataset)

# chisquare
table_gender <- table(diabetes_dataset$gender, diabetes_dataset$diabetes)
chisq.test(table_gender)
```